### 2023-07-09 paper/main.tex
Change format of paper from Markdown to LaTeX

\PassOptionsToPackage{utf8}{inputenc}
\documentclass{bioinfo}

\usepackage{natbib}
%\usepackage[
%backend=biber,
%sorting=ynt
%]{biblatex}

\usepackage{makecell}

\usepackage{floatrow}

\usepackage{comment}

\usepackage{siunitx}

% singlelinecheck=false puts subcaptions on the left
\usepackage[singlelinecheck=false]{subcaption}

\usepackage[usenames,dvipsnames]{xcolor}

%\usepackage{amsthm}
% Says package is already imported?
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}

\usepackage{algorithm2e}
\SetKwRepeat{Do}{do}{while}%

% we squeeze our figures even more together
\captionsetup{belowskip=-2pt}

\SetAlgoLined
\SetKwProg{MyStruct}{Struct}{ contains}{end}

\newcommand{\vocab}{\textbf}
\newcommand{\red}[1]{{\textcolor{Red}{#1}}}
\newcommand{\FIXME}[1]{\red{[FIXME: #1]}}

\usepackage{orcidlink}
\hypersetup{hidelinks}

\usepackage{graphicx}

\def\labelitemi{--}

\copyrightyear{2023} \pubyear{XXXX}

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Genome Analysis}

\begin{document}
\firstpage{1}

\subtitle{Genome Analysis}

\title[FantasticLamp]{FantasticLamp: a genome graph based pipeline for calculating the efficacy of genomic edits}
\author[Schutte \textit{et~al}.]{

Casper~Schutte\,$^{\orcidlink{0000-0003-4245-6842}\text{\sfb 1}}$,
Erik~Garrison\,$^{\orcidlink{0000-0003-3821-631X}\text{\sfb 2}*}$,
Ian~T~Fiddes\,$^{\orcidlink{0000-0002-1580-7443}\text{\sfb 3}}$

}

\address{
$^{\text{\sf 1}}$Department of Bioinformatics and Computational Biology, University of Stellenbosch, Stellenbosch, 7600, Western Cape, South Africa \\
$^{\text{\sf 2}}$Department of Genetics, Genomics and Informatics, University of Tennessee Health Science Center, Memphis, 38163, Tennessee, USA \\
%
}

\corresp{
$^\ast$To whom correspondence should be addressed. \\
% $^\dagger$Contributed equally.\
}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

\abstract{
\textbf{Motivation:}
Accurately calculating the efficacy of genomic edits is crucial to understanding the performance of the editing techniques, in order to optimize methods and improve the success rate of the edits.
Additionally, by understanding the success rate of genomic edits, researchers can identify any potential problems or limitations of the techniques and work to overcome them.
This can help to improve the accuracy and precision of the editing methods, which is essential for many applications, such as creating genetically engineered cells for therapeutic purposes, understanding gene functions, and studying genetic diversity in a population of cells. Current linear alignment methods for quantifying the success of edits suffer from reference bias and fail to capture the complexity involved when dealing with multiple edit states.\\
% Mention current methods and why there is need for better methods.
\textbf{Results:}
We design \textit{FantasticLamp}, an open source genome-graph-based pipeline for calculating the efficacy of genomic edits performed on multiple populations of cells.
It constructs a variation graph from the reference genome and edit template sequences that includes both edited and unedited sequences as paths, and combinations of edits as potential walks through the graph.
It then maps reads from the edited populations onto this graph in order to calculate the coverage of the edited sequences compared to the unedited sequences.
This pipeline aims to provide a quantitative measure of the success of each genomic edit without bias towards the reference or particular single editing constructs.\\
% Reword this paragraph, add detail regarding the process. Mention the successful implementation.
\textbf{Availability:}
\textit{FantasticLamp} is published as free software under the MIT open source license.
Source code and documentation are available at \url{https://github.com/casper-schutte/fantastic-lamp}.
\textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
%\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
}

\maketitle

\section{Introduction}
\label{sec:introduction}
In the field of genome engineering, researchers often perform genomic edits in cells, such as CRISPR/Cas9, TALEN, and ZNF-based systems, to study gene functions and to create new cell lines \citep{gaj2013zfn}.
However, these edits may not always be successful, and it may be challenging to identify and quantify the success rate of these edits in large-scale data sets \citep{guell2014genome}, \citep{van2020delivery}.
Linear approaches that attempt to consider all possible edit states as individual sequences in order to accurately represent complexity can become overly complex themselves and prone to errors \citep{huang2013short}, \citep{mun2021leviosam}.
Additionally, when assessing the success rate of genomic edits, a linear alignment method may introduce reference bias and overlook the complexity of edits, particularly when multiple nearby edits and edit state mixtures are present.

Alternatively, a genome graph approach can provide a more complete and accurate perspective of sequence relationships, while avoiding reference bias and capturing all allele state mixtures present within the targeted edits, even in the case of overlapped edits \citep{eggertsson2017graphtyper}.

Compared to linear alignment methods, genome graphs can provide a more accurate and comprehensive view \citep{garrison2018variation} of the relationships between sequences \citep{paten2017genome} because reads can be mapped to a graph that captures the variation between all the genomes/subsequences, instead of to a single reference. 
The cost of this increased accuracy and the ability to represent complexity is that using genome graphs is more computationally intensive than linear alignments methods \citep{rakocevic2019fast}.

Given these advantages over linear methods, we suggest a pipeline based on genome graph alignment tools would be more effective/better suited for quantifying edit efficacy in multiple populations of edited cells.

% To add:
% Have I covered:
% Enough backround info (what are we doing?)
% Expanding on motivation (why are we doing this?) More detail on why graphs are better?
% Briefly, how we did/will do this?

Here, we present \textit{FantasticLamp}, a simple pipeline that uses genome graph based sequence analysis to calculate/quantify edit efficacy in experiments where multiple edits were performed separately on the same population (wording?). 
% wording? Detail?

\section{Implementation}
\label{sec:implementation}
Here we describe the steps and tools used in creating \textit{FantasticLamp}. 
\subsection{Overview}

Given a reference genome and reads sequenced from multiple edited populations, the pipeline uses a design library CSV file, which contains the intended edits and reference sequences at the intended edit sites, to construct a genome graph.

The genome graph is made up of the reference genome, the intended edit sequences (`homology arms', or `edit homology arms'), and the reference sequences at the edit sites (`reference homology arms').
The construction of the graph involves taking the homology arms and reference homology arms and mapping them to the reference genome using \textit{minimap2} \citep{li2018minimap2}.
%Minimap: 
The inclusion of the reference homology arms is what allows the pipeline to calculate the relative coverage (and thus the edit efficiency) by comparing the edit homology arm coverage to the reference homology arm coverage.

A variation graph is then induced by \textit{seqwish} \citep{garrison2023unbiased} that represents the relationships between the reference genome and both groups of homology arms (reference and edit).
Before the graph is constructed, the nodes are 'chopped' into segments smaller than 256 base pairs and then the graph is sorted using \textit{odgi} \citep{guarracino2022odgi}.
These steps are necessary in order for the graph to then be constructed, again using \textit{odgi}.

The graph is converted into a format that is more efficient and finally indexed, both steps using \textit{vg} \citep{garrison2018variation}.
The pipeline then maps reads from the edited populations to this graph using \textit{vg}, creating a GAF (Gene Annotation Format) file representing the alignment.
A Python script is called to parse the alignment file and calculate the coverage of the homology arms compared to the reference homology arms.
The output of this pipeline is a coverage table in the form of a TSV file, which displays the name of each intended edit sequence along with the homology arm coverage and reference homology arm coverage.
A flow diagram of the steps and files used in the \textit{FantasticLamp} pipeline can be seen in .  
This provides a quantitative measure of the efficacy of each edit.

\subsection{Details}
Go into more detail. At the graph creation step, include a figure showing a locus with edits in Bandage with the counts (BandageNG works now).

Concluding paragraph.
This allows FantasticLamp to simultaneously quantify the efficacy of edits in multiple edited populations, which can be used to test novel editing methods as well as to verify current methods in experiments that involve genomic edits.
By using reads aligned to a graph instead of linear alignment, FantasticLamp can avoid reference bias, making it a useful tool for researchers in the field of genome engineering and other related fields who wish to precisely quantify genome edit states.\\

%We will discuss the tools used, not in extreme detail but enough to justify and explain their use.
% Mention important functions or at least describe how coverage is calculted, etc. Mention problem of shared reads, plasmid integration, etc.

\section{Results}
\label{sec:results}
%This pipeline was initially tested and developed using a data set containing sequencing reads %(mention platform) from experiments where multiple edits (insertions, deletions, and substitutions) were performed on a population of \textit{Saccharomyces cerevisiae} (check this) by Inscripta Inc. (ref?). 
%The initial/preliminary results showed/demonstrated/indicated the pipeline's ability to calculate read coverage (and thus efficacy) of the edits performed. 
%Unfortunately, logistical issues prevented continued use of this data set and an artificial test data set was constructed. This artificial data set was comprised of a tiny reference genome, an accompanying design library CSV file containing reference sequences from the genome and edited sequences, and artificial reads mapping to either the reference or the edit at a particular edit locus. 
%The pipeline accurately calculated the edit coverage and thus provided a quantitative measure of edit efficacy.

The initial development and testing of this pipeline were carried out using a dataset containing sequencing reads from various experiments. These experiments involved multiple alterations - insertions, deletions, and substitutions - on a population of \textit{Saccharomyces cerevisiae}. They were performed in distinct batches by Inscripta Inc. (reference pending).

Preliminary results demonstrated the pipeline's capacity to calculate the read coverage, which acted as a measure of the effectiveness of the implemented edits.
However, due to unforeseen logistical constraints, further usage of this dataset was halted. Consequently, an artificial dataset was constructed for further testing and refinement.

This synthetic dataset encompassed a small reference genome, a related design library CSV, which contained both the reference and edited sequences from the genome, and artificial reads mapping to either the reference or the edit at a specific edit locus. 
Notably, the pipeline was successful in accurately calculating the edit coverage from this dataset, providing a quantitative evaluation of the edit's effectiveness.

%\begin{figure}
%	\centering
%	\fbox{\includegraphics[width=\columnwidth]{flow.pdf}}
%	\caption{Test Caption for Insertion}
%	\label{fig:insert}
%\end{figure}

%\begin{figure}
%	\fbox{\includegraphics{deletion.pdf}}
%	\caption{Test Caption for Deletion}
%	\label{fig:deletion}
%\end{figure}
%
%\begin{figure}
%	\fbox{\includegraphics{substitution.pdf}}
%	\caption{Test Caption for Substitution}
%	\label{fig:substitution}
%\end{figure}

\begin{figure}[htbp]
	\centering
	\begin{minipage}[b]{0.35\columnwidth}
		\includegraphics[width=\textwidth]{panels.pdf}
		%\caption{Test caption for panel figure}
		\label{fig:panel}
	\end{minipage}
	\hfill
	\begin{minipage}[b]{0.64\columnwidth}
		\centering
		\fbox{\includegraphics[width=\textwidth]{flow_test.pdf}}
		%\caption{Test caption for flow diagram}
		\label{fig:flow}
	\end{minipage}
	\caption{(A) Graph representation of a substitution. Edit path shown in green. (B) Graph representation of an insertion. Edit path shown in light blue. (C) Graph representation of a deletion. Edit path shown in dark blue. (D) Diagram depicting the steps, software tools, and file formats use in the \textit{FantasticLamp} pipeline.}
	\label{fig:images}
\end{figure}

\section{Discussion}
\label{sec:discussion}
As novel genome editing processes are developed in the near future, we posit that the need for software tools that can capture the complexity of the edits and the relationship between them will increase.
We show a basic approach for quantifying complex genome editing results, which are difficult to reliably and simply evaluate using a single reference genome that may lead to biased estimates of edit states.
The design library CSV file used in the creation of this pipeline was unique to a specific set of editing experiments performed at Inscripta Inc.
However, the basic format is generic to other editing experiments that users may carry out in the future, with only minimal changes to the script to adjust for difference in formatting of the design library file.
Future users will have to edit the bash script `find\_coverage.sh' such that the correct columns containing the edit- and reference homology arms are extracted.
While the pipeline was tested only on singleplex sequencing data, it is possible to perform analysis using pooled sequencing data, but data availability has limited our ability to thoroughly test this.
In summary, FantasticLamp provides a basic demonstration of the principle of using a variation graph as a reference system to avoid bias when quantifying genomic edits, and stands as a prototype for future work in this space.
\section*{Acknowledgements}
We are grateful to Inscripta Inc for providing the data from editing experiments that was used to create and test this pipeline. We thank Deanna Church for supporting our collaboration.

\section*{Funding}

% Don't think this applies? Check with Erik
\section*{Data availability}
Code and links to data resources used to build this manuscript and its figures can be found in the paper's public repository: \url{https://github.com/casper-schutte/fantastic-lamp}.

\bibliographystyle{natbib}
\bibliography{document}

\end{document}

### 2023-07-09 Test/compare_coverage_read_info.py

@@ -327,6 +327,8 @@ def write_to_tsv(coverage_list):
 
         for i in coverage_list:
             i[0] = i[0][:-1]
+            # This removes the trailing "+" in the names of the paths, which enables BandageNG to correctly label the
+            # nodes when visualising.
             tsv_writer.writerow(i)
 
 

### 2023-07-09 paper/main.tex
Edited small details

@@
 $^{\text{\sf 2}}$Department of Genetics, Genomics and Informatics, University of Tennessee Health Science Center, Memphis, 38163, Tennessee, USA \\
+$^{\text{\sf 3}}$Ian's affiliation goes here \\
 %
@@
 
-Compared to linear alignment methods, genome graphs can provide a more accurate and comprehensive view \citep{garrison2018variation} of the relationships between sequences \citep{paten2017genome} because reads can be mapped to a graph that captures the variation between all the genomes/subsequences, instead of to a single reference.
+Compared to linear alignment methods, genome graphs can provide a more accurate and comprehensive view \citep{garrison2018variation} of the relationships between sequences \citep{paten2017genome} because reads can be mapped to a graph that captures the variation between all the genomes/subsequences, instead of to a single reference, as seen in Figure \ref{fig:images} (A), (B), and (C).
 The cost of this increased accuracy and the ability to represent complexity is that using genome graphs is more computationally intensive than linear alignments methods \citep{rakocevic2019fast}.
@@
 Here we describe the steps and tools used in creating \textit{FantasticLamp}.
-\subsection{Overview}
 
@@
 A Python script is called to parse the alignment file and calculate the coverage of the homology arms compared to the reference homology arms.
-The output of this pipeline is a coverage table in the form of a TSV file, which displays the name of each intended edit sequence along with the homology arm coverage and reference homology arm coverage.
-A flow diagram of the steps and files used in the \textit{FantasticLamp} pipeline can be seen in .  
-This provides a quantitative measure of the efficacy of each edit.
+The output of this pipeline is a coverage table in the form of a TSV file, which displays the name of each intended edit sequence along with the homology arm coverage and reference homology arm coverage providing a quantitative measure of the efficacy of each edit.
 
-\subsection{Details}
-Go into more detail. At the graph creation step, include a figure showing a locus with edits in Bandage with the counts (BandageNG works now).
-
-Concluding paragraph.
-This allows FantasticLamp to simultaneously quantify the efficacy of edits in multiple edited populations, which can be used to test novel editing methods as well as to verify current methods in experiments that involve genomic edits.
-By using reads aligned to a graph instead of linear alignment, FantasticLamp can avoid reference bias, making it a useful tool for researchers in the field of genome engineering and other related fields who wish to precisely quantify genome edit states.\\
+This allows \textit{FantasticLamp} to simultaneously quantify the efficacy of edits in multiple edited populations, which can be used to test novel editing methods as well as to verify current methods in experiments that involve genomic edits.
+By using reads aligned to a graph instead of linear alignment, \textit{FantasticLamp} can avoid reference bias, making it a useful tool for researchers in the field of genome engineering and other related fields who wish to precisely quantify genome edit states.\\
+A flow diagram of the steps and files used in the \textit{FantasticLamp} pipeline can be seen in Figure \ref{fig:images}(D).
 
@@
 Future users will have to edit the bash script `find\_coverage.sh' such that the correct columns containing the edit- and reference homology arms are extracted.
-While the pipeline was tested only on singleplex sequencing data, it is possible to perform analysis using pooled sequencing data, but data availability has limited our ability to thoroughly test this.
-In summary, FantasticLamp provides a basic demonstration of the principle of using a variation graph as a reference system to avoid bias when quantifying genomic edits, and stands as a prototype for future work in this space.
+While the pipeline was tested on artificial data and singleplex sequencing data, it is possible to perform analysis using pooled sequencing data, but data availability has limited our ability to thoroughly test this.
+In summary, \textit{FantasticLamp} provides a basic demonstration of the principle of using a variation graph as a reference system to avoid bias when quantifying genomic edits, and stands as a prototype for future work in this space.
 \section*{Acknowledgements}

### 2023-07-09 paper/main.tex
Edited small details

@@
 Casper~Schutte\,$^{\orcidlink{0000-0003-4245-6842}\text{\sfb 1}}$,
-Erik~Garrison\,$^{\orcidlink{0000-0003-3821-631X}\text{\sfb 2}*}$,
-Ian~T~Fiddes\,$^{\orcidlink{0000-0002-1580-7443}\text{\sfb 3}}$
-
+Ian~T~Fiddes\,$^{\orcidlink{0000-0002-1580-7443}\text{\sfb 2}}$,
+Erik~Garrison\,$^{\orcidlink{0000-0003-3821-631X}\text{\sfb 3}*}$
 }
@@
 $^{\text{\sf 1}}$Department of Bioinformatics and Computational Biology, University of Stellenbosch, Stellenbosch, 7600, Western Cape, South Africa \\
-$^{\text{\sf 2}}$Department of Genetics, Genomics and Informatics, University of Tennessee Health Science Center, Memphis, 38163, Tennessee, USA \\
-$^{\text{\sf 3}}$Ian's affiliation goes here \\
-%
+$^{\text{\sf 2}}$Ian's affiliation goes here \\
+$^{\text{\sf 3}}$Department of Genetics, Genomics and Informatics, University of Tennessee Health Science Center, Memphis, 38163, Tennessee, USA \\
 }

### 2023-07-11 paper/main.tex
add affiliation

@@
 $^{\text{\sf 1}}$Department of Bioinformatics and Computational Biology, University of Stellenbosch, Stellenbosch, 7600, Western Cape, South Africa \\
-$^{\text{\sf 2}}$Ian's affiliation goes here \\
+$^{\text{\sf 2}}$10x Genomics, Pleasanton, CA \\
 $^{\text{\sf 3}}$Department of Genetics, Genomics and Informatics, University of Tennessee Health Science Center, Memphis, 38163, Tennessee, USA \\

### 2023-08-03 paper/main.tex
reworking intro

@@
 
-\section{Introduction}
+\section*{Introduction}
 \label{sec:introduction}
+
+In the field of genome engineering, various methods such as CRISPR/Cas9, TALEN, and ZNF-based systems are employed to create genomic edits in cells for the study of gene functions and the creation of new cell lines \citep{gaj2013zfn}.
+Identifying and quantifying the success rate of these edits in large-scale data sets can be challenging, as unsuccessful edits may occur \citep{guell2014genome}, \citep{van2020delivery}.
+
+Traditional approaches that attempt to enumerate all possible edit states as individual target sequences must confront an exponential number of edit combinations.
+This can increase complexity and introduce errors \citep{huang2013short}, \citep{mun2021leviosam}.
+Moreover, assessing the success rate of genomic edits through a linear alignment method may lead to reference bias, especially when there are multiple nearby edits and mixed edit states.
+
+An alternative to these linear methods is the use of a genome graph approach, which can offer a more comprehensive perspective on sequence relationships, avoiding reference bias, and capturing all allele state mixtures, even in cases of overlapping edits \citep{eggertsson2017graphtyper,Martiniano_2020}.
+Genome graphs provide a more accurate and complete view compared to linear alignment methods by allowing reads to be mapped to a graph that represents a population of sequences and the variations between them, rather than a single reference \citep{garrison2018variation,paten2017genome}.
+
+Given the aforementioned advantages of the genome graph approach, we propose \textit{FantasticLamp}, a pipeline tailored for quantifying edit efficacy in experiments involving multiple edits performed on the same population.
+This pipeline leverages genome graph-based sequence analysis, offering a more suited solution for this context.
+
+Our method involves mapping genomic edits onto structures representable in variation graphs (Figure \ref{fig:images}A-C).
+We subsequently align reads from a population of edited cells to this graph and quantify edit efficiency by comparing the observed edited and unedited allele states (Figure \ref{fig:images}D).
+This integrated approach ensures a more accurate quantification of edit efficacy across multiple populations of edited cells.
+
+\begin{comment}
 In the field of genome engineering, researchers often perform genomic edits in cells, such as CRISPR/Cas9, TALEN, and ZNF-based systems, to study gene functions and to create new cell lines \citep{gaj2013zfn}.
 However, these edits may not always be successful, and it may be challenging to identify and quantify the success rate of these edits in large-scale data sets \citep{guell2014genome}, \citep{van2020delivery}.
-Linear approaches that attempt to consider all possible edit states as individual sequences in order to accurately represent complexity can become overly complex themselves and prone to errors \citep{huang2013short}, \citep{mun2021leviosam}.
-Additionally, when assessing the success rate of genomic edits, a linear alignment method may introduce reference bias and overlook the complexity of edits, particularly when multiple nearby edits and edit state mixtures are present.
+%Because they need to enumerate an exponential number of combinations,
+Approaches that attempt to consider all possible edit states as individual target sequences must consider all combinations of edits, which increases complexity and can lead to errors \citep{huang2013short}, \citep{mun2021leviosam}.
+Additionally, when assessing the success rate of genomic edits, a linear alignment method may introduce reference bias, particularly when multiple nearby edits and edit state mixtures are present.
 
-Alternatively, a genome graph approach can provide a more complete and accurate perspective of sequence relationships, while avoiding reference bias and capturing all allele state mixtures present within the targeted edits, even in the case of overlapped edits \citep{eggertsson2017graphtyper}.
+Alternatively, a genome graph approach can provide a more complete and accurate perspective of sequence relationships, while avoiding reference bias and capturing all allele state mixtures present within the targeted edits, even in the case of overlapped edits \citep{eggertsson2017graphtyper,Martiniano_2020}.
+Compared to linear alignment methods, genome graphs can provide a more accurate and comprehensive view \citep{garrison2018variation} of the relationships between sequences \citep{paten2017genome} because reads can be mapped to a graph that captures a population of sequences and variation between them, instead of to a single reference.
 
-Compared to linear alignment methods, genome graphs can provide a more accurate and comprehensive view \citep{garrison2018variation} of the relationships between sequences \citep{paten2017genome} because reads can be mapped to a graph that captures the variation between all the genomes/subsequences, instead of to a single reference, as seen in Figure \ref{fig:images} (A), (B), and (C).
-The cost of this increased accuracy and the ability to represent complexity is that using genome graphs is more computationally intensive than linear alignments methods \citep{rakocevic2019fast}.
+Given these advantages over linear methods, we suggest a pipeline based on genome graph alignment tools would be better suited for quantifying edit efficacy in multiple populations of edited cells.
+Here, we present \textit{FantasticLamp}, a simple pipeline that uses genome graph based sequence analysis to quantify edit efficacy in experiments where multiple edits were performed separately on the same population.
 
-Given these advantages over linear methods, we suggest a pipeline based on genome graph alignment tools would be more effective/better suited for quantifying edit efficacy in multiple populations of edited cells.
+Our approach maps genomic edits onto structures that we can represent in variation graphs (Figure \ref{fig:images}A-C).
+We then align reads drawn from a population of edited cells to this graph, and quantify edit efficiency by comparing our observations for edited and unedited alleles states.
+\end{comment}
 
-% To add:
-% Have I covered:
-% Enough backround info (what are we doing?)
-% Expanding on motivation (why are we doing this?) More detail on why graphs are better?
-% Briefly, how we did/will do this?
-
-Here, we present \textit{FantasticLamp}, a simple pipeline that uses genome graph based sequence analysis to calculate/quantify edit efficacy in experiments where multiple edits were performed separately on the same population (wording?).
-% wording? Detail?
-
-\section{Implementation}
+\section*{Implementation}
 \label{sec:implementation}
@@
 
-\section{Results}
+\section*{Results}
 \label{sec:results}
@@
 
-\section{Discussion}
+\section*{Discussion}
 \label{sec:discussion}

### 2023-08-03 paper/main.tex
yeah

@@
 In the field of genome engineering, various methods such as CRISPR/Cas9, TALEN, and ZNF-based systems are employed to create genomic edits in cells for the study of gene functions and the creation of new cell lines \citep{gaj2013zfn}.
-Identifying and quantifying the success rate of these edits in large-scale data sets can be challenging, as unsuccessful edits may occur \citep{guell2014genome}, \citep{van2020delivery}.
+Identifying and quantifying the success rate of these edits in large-scale data sets can be challenging due to factors such as variations in editing efficiency, unintended off-target effects, and the complexity of accurately distinguishing between successfully edited and unedited sequences \citep{guell2014genome}, \citep{van2020delivery}.
+%Identifying and quantifying the success rate of these edits in large-scale data sets can be challenging ... \citep{guell2014genome}, \citep{van2020delivery}.
 
@@
 
-\begin{comment}
-In the field of genome engineering, researchers often perform genomic edits in cells, such as CRISPR/Cas9, TALEN, and ZNF-based systems, to study gene functions and to create new cell lines \citep{gaj2013zfn}.
-However, these edits may not always be successful, and it may be challenging to identify and quantify the success rate of these edits in large-scale data sets \citep{guell2014genome}, \citep{van2020delivery}.
-%Because they need to enumerate an exponential number of combinations,
-Approaches that attempt to consider all possible edit states as individual target sequences must consider all combinations of edits, which increases complexity and can lead to errors \citep{huang2013short}, \citep{mun2021leviosam}.
-Additionally, when assessing the success rate of genomic edits, a linear alignment method may introduce reference bias, particularly when multiple nearby edits and edit state mixtures are present.
-
-Alternatively, a genome graph approach can provide a more complete and accurate perspective of sequence relationships, while avoiding reference bias and capturing all allele state mixtures present within the targeted edits, even in the case of overlapped edits \citep{eggertsson2017graphtyper,Martiniano_2020}.
-Compared to linear alignment methods, genome graphs can provide a more accurate and comprehensive view \citep{garrison2018variation} of the relationships between sequences \citep{paten2017genome} because reads can be mapped to a graph that captures a population of sequences and variation between them, instead of to a single reference.
-
-Given these advantages over linear methods, we suggest a pipeline based on genome graph alignment tools would be better suited for quantifying edit efficacy in multiple populations of edited cells.
-Here, we present \textit{FantasticLamp}, a simple pipeline that uses genome graph based sequence analysis to quantify edit efficacy in experiments where multiple edits were performed separately on the same population.
-
-Our approach maps genomic edits onto structures that we can represent in variation graphs (Figure \ref{fig:images}A-C).
-We then align reads drawn from a population of edited cells to this graph, and quantify edit efficiency by comparing our observations for edited and unedited alleles states.
-\end{comment}
-
 \section*{Implementation}
@@
 
-Given a reference genome and reads sequenced from multiple edited populations, the pipeline uses a design library CSV file, which contains the intended edits and reference sequences at the intended edit sites, to construct a genome graph.
+Given a reference genome and reads sequenced from multiple edited populations, the pipeline uses a design library file---a table which contains the intended edits and reference sequences at the intended edit sites---to construct a genome graph.
 

### 2023-08-03 paper/main.tex
ok

@@
 \label{sec:implementation}
-Here we describe the steps and tools used in creating \textit{FantasticLamp}.
+\section*{Implementation}
+\label{sec:implementation}
+Here we describe the steps and tools utilized in creating \textit{FantasticLamp}.
 
-Given a reference genome and reads sequenced from multiple edited populations, the pipeline uses a design library file---a table which contains the intended edits and reference sequences at the intended edit sites---to construct a genome graph.
+Given a reference genome and reads sequenced from multiple edited populations, the pipeline employs a design library file---a table containing the intended edits and reference sequences at the intended edit sites---to construct a genome graph.
 
-The genome graph is made up of the reference genome, the intended edit sequences (`homology arms', or `edit homology arms'), and the reference sequences at the edit sites (`reference homology arms').
-The construction of the graph involves taking the homology arms and reference homology arms and mapping them to the reference genome using \textit{minimap2} \citep{li2018minimap2}.
-%Minimap: 
-The inclusion of the reference homology arms is what allows the pipeline to calculate the relative coverage (and thus the edit efficiency) by comparing the edit homology arm coverage to the reference homology arm coverage.
+The genome graph consists of the reference genome, the intended edit sequences (referred to as `homology arms' or `edit homology arms'), and the reference sequences at the edit sites (termed `reference homology arms').
+The core structure of the graph is an alignment of homology and reference homology arms onto the reference genome, which we complet using \textit{minimap2} \citep{li2018minimap2}.
+The inclusion of the reference homology arms enables the pipeline to calculate the relative coverage, thereby determining the edit efficiency by comparing the coverage of edit homology arms to that of the reference homology arms.
 
-A variation graph is then induced by \textit{seqwish} \citep{garrison2023unbiased} that represents the relationships between the reference genome and both groups of homology arms (reference and edit).
-Before the graph is constructed, the nodes are 'chopped' into segments smaller than 256 base pairs and then the graph is sorted using \textit{odgi} \citep{guarracino2022odgi}.
-These steps are necessary in order for the graph to then be constructed, again using \textit{odgi}.
+Then, we construct the graph itself.
+A variation graph, representing the relationships between the reference genome and both groups of homology arms (reference and edit), is then induced by \textit{seqwish} \citep{garrison2023unbiased}, which combines the input sequences and the alignments into a graph (in GFAv1 format).
+After the graph's construction, the nodes are 'chopped' into segments smaller than 256 base pairs to enable efficient indexing and read mapping, and the graph is sorted using \textit{odgi} \citep{guarracino2022odgi}.
 
-The graph is converted into a format that is more efficient and finally indexed, both steps using \textit{vg} \citep{garrison2018variation}.
-The pipeline then maps reads from the edited populations to this graph using \textit{vg}, creating a GAF (Gene Annotation Format) file representing the alignment.
-A Python script is called to parse the alignment file and calculate the coverage of the homology arms compared to the reference homology arms.
-The output of this pipeline is a coverage table in the form of a TSV file, which displays the name of each intended edit sequence along with the homology arm coverage and reference homology arm coverage providing a quantitative measure of the efficacy of each edit.
+The pipeline then maps reads from the edited populations to this graph using \textit{vg} \citep{garrison2018variation}, generating a GAF (Gene Annotation Format) file to represent the alignment.
+A Python script parses the alignment file to calculate the coverage of the homology arms relative to the reference homology arms.
+We output a coverage table (in TSV), displaying each intended edit sequence alongside the corresponding homology arm coverage and reference homology arm coverage, thus quantifying the efficacy of each edit.
 
-This allows \textit{FantasticLamp} to simultaneously quantify the efficacy of edits in multiple edited populations, which can be used to test novel editing methods as well as to verify current methods in experiments that involve genomic edits.
-By using reads aligned to a graph instead of linear alignment, \textit{FantasticLamp} can avoid reference bias, making it a useful tool for researchers in the field of genome engineering and other related fields who wish to precisely quantify genome edit states.\\
-A flow diagram of the steps and files used in the \textit{FantasticLamp} pipeline can be seen in Figure \ref{fig:images}(D).
+\textit{FantasticLamp} thus enables the simultaneous quantification of edit efficacy across multiple edited populations, serving as a valuable tool for both the testing of novel editing methods and the verification of current methods in experiments involving genomic edits.
+By aligning reads to a graph instead of employing linear alignment, \textit{FantasticLamp} circumvents reference bias and will allow for precise quantification of genome edit states even in the case of contexts where many simultaneous edits are applied in close proximity.
+%in genome engineering and related fields who seek precise quantification of genome edit states. %\\
+A flow diagram detailing the steps and files used in the \textit{FantasticLamp} pipeline is presented in Figure \ref{fig:images}D.
 

### 2023-08-03 paper/main.tex
...

@@
 The initial development and testing of this pipeline were carried out using a dataset containing sequencing reads from various experiments.
-These experiments involved multiple alterations - insertions, deletions, and substitutions - on a population of \textit{Saccharomyces cerevisiae}.
-They were performed in distinct batches by Inscripta Inc.
-(reference pending).
+These experiments involved multiple alterations---insertions, deletions, and substitutions---on a population of \textit{Saccharomyces cerevisiae}.
+They were performed in distinct batches by Inscripta Inc. \cite{gander2021simultaneous}.
 
 Preliminary results demonstrated the pipeline's capacity to calculate the read coverage, which acted as a measure of the effectiveness of the implemented edits.
-However, due to unforeseen logistical constraints, further usage of this dataset was halted. Consequently, an artificial dataset was constructed for further testing and refinement.
+However, due to unforeseen logistical constraints, further usage of this dataset was halted.
+Consequently, an artificial dataset was constructed for further testing and refinement.
 
-This synthetic dataset encompassed a small reference genome, a related design library CSV, which contained both the reference and edited sequences from the genome, and artificial reads mapping to either the reference or the edit at a specific edit locus.
-Notably, the pipeline was successful in accurately calculating the edit coverage from this dataset, providing a quantitative evaluation of the edit's effectiveness.
+This synthetic dataset encompasses a small reference genome, a related design library CSV, which contain both the reference and edited sequences from the genome, and artificial reads mapping to either the reference or the edit at a specific edit locus.
+\textit{FantasticLamp} accurately calculated the edit coverage from this dataset, providing a quantitative evaluation of the edit's effectiveness.
+We provide this as a standard test in our code repository.
 

### 2023-08-03 paper/main.tex
fun

@@
 
-\access{Advance Access Publication Date: Day Month Year}
-\appnotes{Genome Analysis}
+%\access{Advance Access Publication Date: Day Month Year}
+%\appnotes{Genome Analysis}
 
@@
 
-\subtitle{Genome Analysis}
+%\subtitle{Genome Analysis}
 
@@
 These experiments involved multiple alterations---insertions, deletions, and substitutions---on a population of \textit{Saccharomyces cerevisiae}.
-They were performed in distinct batches by Inscripta Inc. \cite{gander2021simultaneous}.
+They were performed in distinct batches by Inscripta Inc. \citep{gander2021simultaneous}.
 
@@
 \label{sec:discussion}
+\textit{FantasticLamp} demonstrates using a variation graph to quantify complex genome edits.
+We consider this work as an example of how to apply unbiased edit quantification to editing experiments.
+However, much work is left to be done to use these approaches at scale.
+As an emerging field, commodity high-throughput genome editing is still rare and we lack accepted standards for experimental design, so our implementation's inputs and outputs are necessarily bespoke.
+While we demonstrate basic functionality of \textit{FantasticLamp} on artificial and singleplex data, it could be expanded to analyze multiplexed pooled data.
+This pipeline provides a starting point to avoid bias when quantifying edits that can be built upon as editing technologies advance.
+Potential improvements include handling more complex structural variants, incorporating multiplexed data, estimating off-target effects, optimizing runtime for large genomes, and interfacing for non-expert users.
+In summary, \textit{FantasticLamp} provides a basic demonstration of the principle of using a variation graph as a reference system to avoid bias when quantifying genomic edits.
+
+\begin{comment}
+\section*{Discussion}
+\label{sec:discussion}
 As novel genome editing processes are developed in the near future, we posit that the need for software tools that can capture the complexity of the edits and the relationship between them will increase.
@@
 In summary, \textit{FantasticLamp} provides a basic demonstration of the principle of using a variation graph as a reference system to avoid bias when quantifying genomic edits, and stands as a prototype for future work in this space.
+\end{comment}
+
 \section*{Acknowledgements}

### 2023-08-03 paper/main.tex
ok but where are the experiments

@@
 This can help to improve the accuracy and precision of the editing methods, which is essential for many applications, such as creating genetically engineered cells for therapeutic purposes, understanding gene functions, and studying genetic diversity in a population of cells.
-Current linear alignment methods for quantifying the success of edits suffer from reference bias and fail to capture the complexity involved when dealing with multiple edit states.\\
+Current linear alignment methods for quantifying the success of edits exhibit reference bias and cannot model complex mixed edit states.\\
 % Mention current methods and why there is need for better methods.
@@
 A Python script parses the alignment file to calculate the coverage of the homology arms relative to the reference homology arms.
-We output a coverage table (in TSV), displaying each intended edit sequence alongside the corresponding homology arm coverage and reference homology arm coverage, thus quantifying the efficacy of each edit.
+We output a coverage table (in TSV format), displaying each intended edit sequence alongside the corresponding homology arm coverage and reference homology arm coverage, thereby quantifying the efficacy of each edit.
 
@@
 
-Preliminary results demonstrated the pipeline's capacity to calculate the read coverage, which acted as a measure of the effectiveness of the implemented edits.
+We found that the pipeline captured expected edit states.
 However, due to unforeseen logistical constraints, further usage of this dataset was halted.
@@
 However, much work is left to be done to use these approaches at scale.
-As an emerging field, commodity high-throughput genome editing is still rare and we lack accepted standards for experimental design, so our implementation's inputs and outputs are necessarily bespoke.
+%As an emerging field, commodity high-throughput genome editing is still rare and we lack accepted standards for experimental design, so our implementation's inputs and outputs are necessarily bespoke.
+As an emerging field, commodity high-throughput genome editing is still uncommon, and accepted standards for experimental design are lacking.
+Therefore, the inputs and outputs of our implementation are necessarily customized.
+
+This pipeline provides a starting point to avoid bias when quantifying edits that can be built upon as editing technologies advance.
 While we demonstrate basic functionality of \textit{FantasticLamp} on artificial and singleplex data, it could be expanded to analyze multiplexed pooled data.
-This pipeline provides a starting point to avoid bias when quantifying edits that can be built upon as editing technologies advance.
-Potential improvements include handling more complex structural variants, incorporating multiplexed data, estimating off-target effects, optimizing runtime for large genomes, and interfacing for non-expert users.
+Other potential improvements include handling more complex structural variants, incorporating multiplexed data, estimating off-target effects, and optimizing runtime for large genomes.
 In summary, \textit{FantasticLamp} provides a basic demonstration of the principle of using a variation graph as a reference system to avoid bias when quantifying genomic edits.

### 2024-01-25 paper/main.tex
oops
Updated manuscript

@@
 \label{sec:implementation}
-\section*{Implementation}
-\label{sec:implementation}
 Here we describe the steps and tools utilized in creating \textit{FantasticLamp}.
@@
 We provide this as a standard test in our code repository.
+
+For further validation, we created a much larger validation test by taking a full yeast genome \citep{ncbis288c2023} and introducing a single edit (ranging from 20 - 25bp) at a specific locus, which we then simulated reads from.
+We created 100 of these edited genomes, each with a different edit at a random locus, and simulated 10x coverage, paired-end reads from each genome.
+We then ran \textit{FantasticLamp} on two sets of these simulated reads; one set with a base error rate of 0.5\%, and one with a base error rate of 1\%.
+In both of these validation tests, \textit{FantasticLamp} was able to identify the correct edit state for 97 out of 100 edits.
+The scripts used to generate these genomes and reads (as well as to perform the actual validation tests) are available in the code repository.
 
@@
 However, much work is left to be done to use these approaches at scale.
-%As an emerging field, commodity high-throughput genome editing is still rare and we lack accepted standards for experimental design, so our implementation's inputs and outputs are necessarily bespoke.
+
 As an emerging field, commodity high-throughput genome editing is still uncommon, and accepted standards for experimental design are lacking.
@@
 This pipeline provides a starting point to avoid bias when quantifying edits that can be built upon as editing technologies advance.
-While we demonstrate basic functionality of \textit{FantasticLamp} on artificial and singleplex data, it could be expanded to analyze multiplexed pooled data.
+In this paper, we demonstrate basic functionality of \textit{FantasticLamp} on artificial and singleplex data. 
+Using a large validation test, we confirmed the robustness and accuracy of the approaches outlined in this paper.
+Moreover, the potential of \textit{FantasticLamp} extends beyond singleplex data, as it could be further developed to analyze multiplexed pooled data.
 Other potential improvements include handling more complex structural variants, incorporating multiplexed data, estimating off-target effects, and optimizing runtime for large genomes.
-In summary, \textit{FantasticLamp} provides a basic demonstration of the principle of using a variation graph as a reference system to avoid bias when quantifying genomic edits.
+In summary, \textit{FantasticLamp} provides a basic demonstration of the principle of using a variation graph as a reference system to avoid bias when accurately quantifying genomic edits.
 

### 2024-02-01 paper/main.tex
Updated README

@@
 \section*{Data availability}
-Code and links to data resources used to build this manuscript and its figures can be found in the paper's public repository: \url{https://github.com/casper-schutte/fantastic-lamp}.
+The scripts that comprise the \textit{FantasticLamp} pipeline, as well as the code and links to data resources used to build this manuscript and its figures, can be found in the paper's public repository: \url{https://github.com/casper-schutte/fantastic-lamp}.
 

### 2024-02-01 paper/main.tex
minor edits

@@
 
-\title[FantasticLamp]{FantasticLamp: a genome graph based pipeline for calculating the efficacy of genomic edits}
+\title[FantasticLamp]{FantasticLamp: quantifying genomic edits with genome variation graphs}
 \author[Schutte \textit{et~al}.]{
@@
 An alternative to these linear methods is the use of a genome graph approach, which can offer a more comprehensive perspective on sequence relationships, avoiding reference bias, and capturing all allele state mixtures, even in cases of overlapping edits \citep{eggertsson2017graphtyper,Martiniano_2020}.
-Genome graphs provide a more accurate and complete view compared to linear alignment methods by allowing reads to be mapped to a graph that represents a population of sequences and the variations between them, rather than a single reference \citep{garrison2018variation,paten2017genome}.
+Genome graphs provide a more accurate and complete view compared to linear alignment methods by allowing reads to be mapped to a graph that represents a population of sequences and the variations between them, rather than a single reference \citep{garrison2018variation,paten2017genome,Eizenga_2020}.
 
@@
 
-The genome graph consists of the reference genome, the intended edit sequences (referred to as `homology arms' or `edit homology arms'), and the reference sequences at the edit sites (termed `reference homology arms').
-The core structure of the graph is an alignment of homology and reference homology arms onto the reference genome, which we complet using \textit{minimap2} \citep{li2018minimap2}.
+The genome graph consists of the reference genome, the intended edit sequences (referred to as ``homology arms'' or ``edit homology arms''), and the reference sequences at the edit sites (termed ``reference homology arms'').
+The core structure of the graph is an alignment of homology and reference homology arms onto the reference genome, which we complete using \textit{minimap2} \citep{li2018minimap2}.
 The inclusion of the reference homology arms enables the pipeline to calculate the relative coverage, thereby determining the edit efficiency by comparing the coverage of edit homology arms to that of the reference homology arms.
@@
 
-For further validation, we created a much larger validation test by taking a full yeast genome \citep{ncbis288c2023} and introducing a single edit (ranging from 20 - 25bp) at a specific locus, which we then simulated reads from.
+For further validation, we created a much larger validation test by taking a full yeast genome (GCF\_000146045.2) and introducing a single edit (ranging from 20 - 25bp) at a specific locus, which we then simulated reads from.
 We created 100 of these edited genomes, each with a different edit at a random locus, and simulated 10x coverage, paired-end reads from each genome.

### 2024-02-02 paper/main.tex
funding and conflict of interest

@@
 \section*{Funding}
+The authors gratefully acknowledge support from National Institutes of Health/NIDA U01DA047638 (E.G.), National Institutes of Health/NIGMS R01GM123489 (E.G.), and NSF PPoSS Award \#2118709 (E.G.), and the Center for Integrative and Translational Genomics (E.G.
+and C.S.).
 
@@
 
+\section*{Conflict of interest}
+The authors declare that they have no conflicts of interest.
+
 \bibliographystyle{natbib}
